---
title: Analyzing polymapR's example using MAPpoly 
author: "Marcelo Mollinari"
date: "`r Sys.Date()`"
output:
 html_document:
   highlight: tango
   keep_md: yes
   theme: united
   toc: yes
   toc_float:
     collapsed: no
 md_document:
   variant: markdown_github
 pdf_document:
   toc: yes
   toc_depth: '3'
   highlight: tango
linestretch: 1.2
#bibliography: biblio.bib
#output: rmarkdown::html_vignette
#vignette: >
#  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
#load("~/repos/tutorials/polymapR_example/mapping_for_polymapR_example.rda")
load("~/repos/tutorials/polymapR_example/polymapR_exemple.rda")
require(ggplot2)
```

# Context

Recently, [Bourke et al. (2018)](https://academic.oup.com/bioinformatics/advance-article/doi/10.1093/bioinformatics/bty371/4990829) released an R package for constructing genetic maps in species with ploidy level up to 6 called [polymapR](https://cran.r-project.org/web/packages/polymapR). The software uses pairwise recombination fractions coupled with the MDS algorithm [(Preddy and Hackett, 2016)](https://link.springer.com/article/10.1007%2Fs00122-016-2761-8) to assemble linkage groups, order, and phase genetic markers. The final genetic distance between markers is obtained using the projection of the MDS configuration onto a principal curve, as also proposed by [Preddy and Hackett (2016)](https://link.springer.com/article/10.1007%2Fs00122-016-2761-8). Although the method is fast and provides reliable results under certain circumstances (i.e., presence of informative markers), we noticed that, in order to avoid data with high levels of noise, the analytic procedure described in [polymapR's tutorial](https://cran.r-project.org/web/packages/polymapR/vignettes/Vignette_polymapR.html) removes a considerable portion of the markers. Here, we use our [MAPpoly software](https://github.com/mmollina/MAPpoly) to construct a genetic map using the same [data set](https://cran.r-project.org/web/packages/polymapR/vignettes/Vignette_polymapR.html#sect6) presented in polymapR's tutorial and compare the results. 

# Overview of our strategy

We assembled the linkage groups (LGs) using the full recombination fraction matrix and standard UPGMA procedures. Within each linkage group,  we also relied on MDS algorithm [(Preddy and Hackett, 2016)](https://link.springer.com/article/10.1007%2Fs00122-016-2761-8) implementation to order markers. For a given order, we applied our sequential algorithm [(Mollinari and Garcia, 2018)](https://www.biorxiv.org/content/early/2018/09/12/415232) which uses pairwise recombination fraction as the first source of information to sequentially position the allelic variants in specific homologous chromosomes. For situations where pairwise analysis has limited power, it relies on the likelihood obtained through a hidden Markov model (HMM). Once all markers were positioned, the final map was reconstructed using the HMM multilocus algorithm. Thus, differently from `polymapR`, our method does not rely on single dose markers to build a homologous framework. 

Using our procedure we were able to increase the number of phased markers in the final map by about 2.56 fold. Also, the recombination fractions were multilocus estimated (i.e., using the maximum likelihood of the HMM). Finally, we included a global genotype error rate of 0.05 in the HMM emission function to reestimate the final map. In real situations, this error could be marker/offspring specific and is obtained using several genotype-calling methods available, such as [ClusterCall](https://potatobreeding.cals.wisc.edu/software/), [fitTetra](https://www.wur.nl/en/show/Software-fitTetra.htm), [SuperMASSA](https://bitbucket.org/orserang/supermassa) and [updog](https://github.com/dcgerard/updog). The final length of the maps was somewhat different due to the different estimation procedures: `MAPpoly` yield a longer map when compared to `polymapR` (MAPpoly: 101.76 cM/lg and polymapR: 90.2 cM/lg, in average). We compared our phasing results with `polymapR`'s result and, for the markers contained in both maps, the results were precisely the same for all linkage groups. However, since our method does not rely on single dose markers to perform the phasing, we could include 2514 markers in the final map (from a total of 2873 markers, 256 % more than `polymapR` result, which included 982 markers). In terms of performance, since our program use multilocus analysis and model genotype errors, it takes about 30 minutes to automatically build the map. On the other hand, `polymapR` uses only two point analysis, making the recombination fraction computation faster.
Thus, using the script from polymapR’s tutorial, the map construction takes less than one minute. 

# Conclusion
Depending on the amount of information available in the data set, `polymapR` can efficiently produce good results. However, a high level of "noise-to-signal" ratio is expected in polyploid genotypic data sets. `MAPpoly`'s approach extract more information from the data set and is capable of modeling genotype errors due to its multilocus nature. Moreover, it does not depend on single dose markers to provide a phasing framework. It is worthwhile to mention that the construction of complex polyploid maps could be challenging if the adequate computational power is not available. In our experience, we were able to construct an ultra-saturated map of the hexaploid sweetpotato containing more than 38,000 SNPs, where 55% where simplex and 45% multiplex. 

# Map construction

## Loading `polymapR` results

There are two map results in `polymapR`'s tutorial: the first one is called "integrated map"", which contains markers from both parents, but not totally phased. The second one is the "phased map", with less but phased markers. 

```{r load_polymapR}
require(polymapR)
load("~/repos/tutorials/polymapR_example/polymapR_exemple.rda")
plot_map(maplist = integrated.maplist)
plot_map(maplist = phased.maplist)
```

Summary for both maps:

```{r polymapr_maps_summary_load, echo=FALSE}
polymapR.integrated.maps<-sapply(integrated.maplist, function(x) c(nrow(x), 
                                                           round(max(x$position),1),
                                                           round(max(x$position)/nrow(x),2)))
polymapR.phased.maps<-sapply(phased.maplist, function(x) c(nrow(x), 
                                                           round(max(x$position),1),
                                                           round(max(x$position)/nrow(x),2)))
rownames(polymapR.integrated.maps) <- rownames(polymapR.phased.maps) <- c("n.markers", "length", "cM/mrk")
```

The number of markers, length and markers density are presented bellow:

```{r polymapr_maps_summary}
polymapR.integrated.maps
polymapR.phased.maps
```

## Map construction using `MAPPoly`


This is a very brief tutorial. For a detailed procedure to build a map using `MAPPoly`, please refer to [this tutorial](https://mmollina.github.io/MAPPoly/). 

### Loading and converting data

Let us load `MAPPoly` and convert the data set used in `polymapR` tutorial. We will use the `"segregating_data"` data set which was already filtered for no-segregating markers. Also, all the segregation types were converted in their simplest form. This is not a requirement for `MAPPoly` but we are using the same data set to compare results

```{r load_mappoly}
require(mappoly)
source("~/repos/tutorials/polymapR_example/polymapR_to_mappoly.R")
data("segregating_data")
```

Summary of both data sets:

```{r dataset}
tetra.data<-polymapR_to_mappoly(segregating_data)
print(tetra.data, detailed = TRUE)
```

### Recombination fraction estimation

Now, let us compute the recombination fraction between all markers in the data set. 

```{r rf_estimation, eval=TRUE}
s<-make_seq_mappoly(tetra.data, arg = "all")
counts.web<-cache_counts_twopt(s, get.from.web = TRUE)
all.pairs<-est_pairwise_rf(input.seq = s,
                           count.cache = counts.web,
                           n.clusters = 16, #toke approx. 9 minutes using 24 CPUs in BRC cluster
                           verbose=TRUE) 
mat.full<-rf_list_to_matrix(input.twopt=all.pairs)
mat.full
```

### Grouping

Assigning markers to linkage groups is an interactive task in MAPpoly. Since we expect five linkage groups, we tried to use 5 as the expected number. However, due to some markers that were not clearly positioned in any group, we divided the resulting dendrogram into a larger number of groups and picked the densest five clusters.
```{r grouping, warning=FALSE}
lgs<-group_mappoly(input.mat = mat.full,
                   input.seq = s,
                   expected.groups = 11,
                   comp.mat = FALSE, 
                   inter = FALSE)
lgs
plot(lgs)
## Selecting groups
LGS<-lapply(c(1,2,4,5,6), function(x, lgs) make_seq_mappoly(lgs, x), lgs)
```

### Filtering and ordering markers within groups

Next, we use the function `rf_snp_filter` to remove markers that do not meet a LOD and recombination fraction criteria for at least 10% of the pairwise markers combinations within each linkage group. After that, we obtain the recombination fraction matrices for all groups and use the MDS algorithm to order the markers.

```{r mds, echo=FALSE, eval=TRUE}
P<-lapply(LGS, make_pairs_mappoly, input.twopt = all.pairs)
LGS.filt<-lapply(P, rf_snp_filter, thresh.LOD.ph = 5, 
                 thresh.LOD.rf = 5, thresh.rf = 0.15, 
                 thresh.perc = 0.10)
P.filt<-lapply(LGS.filt, make_pairs_mappoly, input.twopt = all.pairs)
M.filt<-lapply(P.filt, rf_list_to_matrix)
cl <- parallel::makeCluster(5)
parallel::clusterEvalQ(cl, require(mappoly))
MDS.ord <- parallel::parLapply(cl,
                               M.filt,
                               mds_mappoly)
parallel::stopCluster(cl)
MDS.seq<-lapply(MDS.ord, make_seq_mappoly)
```

Now, let us plot the reordered recombination fraction 
```{r mds_plot}
op <- par(mfrow = c(2, 3), pty = "s")
for(i in 1:5)
  plot(M.filt[[i]], ord = MDS.seq[[i]]$seq.mrk.names, 
       main.text = paste0("LG", i), index = FALSE)
par(op)
```

Now, the full recombination fraction matrix
```{r full_rf_mat, fig.width=10, fig.height=10}
o<-unlist(sapply(MDS.seq, function(x) x$seq.mrk.names))
plot(mat.full, ord = o)
```

### Phasing and multilocus reconstruction

Given the MDS order, we estimate the phase and the recombination fraction between all adjacent markers. After that, we reestimate the map using the function `est_full_hmm_with_global_error`, which includes a global genotype error in the HMM model.

```{r phasing_and_reestimate, eval=TRUE}
## function for parallel computation
my.phase.func<-function(X)
{
  return(est_rf_hmm_sequential(input.seq = X[[1]],
                               thres.twopt = 10,
                               thres.hmm = 10,
                               extend.tail = 20,
                               twopt = X[[2]],
                               verbose = TRUE,
                               tol = 10e-3,
                               tol.final = 10e-4,
                               phase.number.limit = 20,
                               sub.map.size.diff.limit = 4,
                               info.tail = TRUE,
                               reestimate.single.ph.configuration = TRUE, 
                               high.prec = FALSE))
  
}
## list with ordered sequences and pairwise recombination fractions
X<-vector("list", 5)
for(i in 1:5)
  X[[i]]<-list(MDS.seq[[i]], P.filt[[i]])

## Running phasing algorithm in 5 CPUs
system.time({
  cl <- parallel::makeCluster(5)
  parallel::clusterEvalQ(cl, require(mappoly))
  parallel::clusterExport(cl, "tetra.data")
  MAPS <- parallel::parLapply(cl,X,my.phase.func)
  MAPs.res <- parallel::parLapply(cl, MAPS, est_full_hmm_with_global_error, 
                                  error = 0.05, tol = 10e-4)
  parallel::stopCluster(cl)
})
```

### Comapring maps

Now, let us compare the results from both programs. 
```{r }
id<-c(4,3,1,2,5) ##This is to match the linkage group ordering in polymapR tutorial
nm<-pal<-comp.PHQ<-comp.PHP<-vector("list", 5)
for(i in 1:5)
{
  YP<-phased.maplist[[id[i]]][,3:6]
  rownames(YP)<-phased.maplist[[id[i]]][,1]
  YQ<-phased.maplist[[id[i]]][,7:10]
  rownames(YQ)<-phased.maplist[[id[i]]][,1]
  mrk.names<-tetra.data$mrk.names[MAPS[[i]]$maps[[1]]$seq.num]
  nm[[i]]<-intersect(rownames(YQ), mrk.names)
  AP<-MAPS[[i]]$maps[[1]]$seq.ph$P
  AQ<-MAPS[[i]]$maps[[1]]$seq.ph$Q
  names(AQ)<-names(AP)<-mrk.names
  BP<-ph_matrix_to_list(YP)
  names(BP)<-rownames(YP)
  BQ<-ph_matrix_to_list(YQ)
  names(BQ)<-rownames(YQ)
  comp.PHP[[i]]<-compare_haplotypes(m = 4, h1 = AP[nm[[i]]], h2 = BP[nm[[i]]]) 
  comp.PHQ[[i]]<-compare_haplotypes(m = 4, h1 = AQ[nm[[i]]], h2 = BQ[nm[[i]]]) 
  pal[[i]]<-mrk.names%in%rownames(YQ)+1
}

res<-rbind(sapply(phased.maplist, nrow),
           sapply(MAPS, function(x) x$info$n.mrk),
           sapply(phased.maplist, function(x) round(max(x$position,1))),
           sapply(MAPs.res, function(x) round(sum(imf_h(x$maps[[1]]$seq.rf)),1)),
           sapply(nm, length),
           sapply(comp.PHP, function(x) x$is.same.haplo),
           sapply(comp.PHQ, function(x) x$is.same.haplo))
dimnames(res)<-list(c("n.mrk.polymapR",  "n.mrk.MAPpoly" , "polymapR.length", "MAPPoly.length","intersect", "is.same.phase.P1", "is.same.phase.P2"), c("LG1","LG2","LG3","LG4","LG5"))
```

The columns in the following table present the number of markers assigned to each LG (rows) using `polymapR` and `MAPpoly`, respectively, followed by the length of the resulting maps, the number of markers shared between the maps and whether the estimated linkage phase was the same within the shared markers (1) or not (0). 
```{r}
t(res)
```

Comparing number of markers assigned to each LG using both methods

```{r, echo=FALSE}
df1<-reshape::melt(res[1:2,])
colnames(df1)<-c("method", "LG", "n.markers")
df1$method<-substring(df1$method,7)
ggplot(data=df1, aes(x=LG, y=n.markers, fill=method)) +
  geom_bar(stat="identity", position=position_dodge())+
  geom_text(aes(label=n.markers), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()  +
  xlab("linkage groups")  + ylab("number of markers")
```

Comparing the map length of each LG using both methods

```{r, echo=FALSE}
df2<-reshape::melt(res[3:4,])
colnames(df2)<-c("method", "LG", "map.length")
df2$method<-df1$method
ggplot(data=df2, aes(x=LG, y=map.length, fill=method)) +
  geom_bar(stat="identity", position=position_dodge())+
  geom_text(aes(label=map.length), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal() +
  xlab("linkage groups")  + ylab("map length")
```
Next, we compare the likelihoods of the maps produced by both programs using only the shared markers. Since the phase is the same in both cases, here we are evaluating the order obtained using MDS. Notice that, the multilocus likelihood is available only in `MAPpoly`. Thus we reestimated the maps produced by `polymapR` with our HMM procedure.
```{r check, echo=TRUE, eval=TRUE}
a<-b<-P.MAPs.temp<-M.MAPs.temp<-vector("list", 5)
for(i in 1:5)
{
  a[[i]]<-integrated.maplist[[id[i]]][,1:2]
  b[[i]]<-data.frame(marker = tetra.data$mrk.names[MAPS[[i]]$maps[[1]]$seq.num],
                     position = cumsum(mappoly::imf_h(c(0, MAPS[[i]]$maps[[1]]$seq.rf))))
  s.compP<-mappoly::make_seq_mappoly(tetra.data, arg = intersect(a[[i]]$marker, b[[i]]$marker))
  s.compM<-mappoly::make_seq_mappoly(tetra.data, arg = intersect(b[[i]]$marker, a[[i]]$marker))
  oP<-match(s.compP$seq.num, MAPs.res[[i]]$maps[[1]]$seq.num)
  oM<-match(s.compM$seq.num, MAPs.res[[i]]$maps[[1]]$seq.num)
  P.MAPs.temp[[i]]<-mappoly::get_submap(MAPs.res[[i]], mrk.pos = oP, 
                                        reestimate.rf = TRUE, 
                                        tol.final = 10e-2, verbose = FALSE)
  M.MAPs.temp[[i]]<-mappoly::get_submap(MAPs.res[[i]], mrk.pos = oM,  
                                        reestimate.rf = TRUE, 
                                        tol.final = 10e-2, verbose = FALSE)
}
```

Except from linkage group #2, based on the multilocus likelihood, the order of the shared markers was superior when using `MAPpoly`. 

```{r error, echo=FALSE}
cl <- parallel::makeCluster(5)
parallel::clusterEvalQ(cl, require(mappoly))
parallel::clusterExport(cl, "tetra.data")
like.P.MAPs <- parallel::parLapply(cl, P.MAPs.temp, est_full_hmm_with_global_error, 
                                   error = 0.05, tol = 10e-4)
like.M.MAPs <- parallel::parLapply(cl, M.MAPs.temp, est_full_hmm_with_global_error, 
                                   error = 0.05, tol = 10e-4)
parallel::stopCluster(cl)
```

```{r, echo=FALSE}
x<-NULL
for(i in 1:5)
  x<-rbind(x, c(like.P.MAPs[[i]]$maps[[1]]$loglike, like.M.MAPs[[i]]$maps[[1]]$loglike))
colnames(x)<-c("polymapR", "MAPPoly")
t(apply(x, 1, function(x) max(x)-x))
```

Now, let us compare both maps. Maps to the left were produced by `polymapR`; maps to the left were produced by `MAPpoly`. Shared markers (red dots) are linked by black lines. Private markers are represented by blue dots. 

```{r, echo=FALSE, eval=TRUE, fig.width=10, fig.height=10}
b[[2]]<-data.frame(marker = tetra.data$mrk.names[rev(MAPS[[2]]$maps[[1]]$seq.num)],
                   position = cumsum(mappoly::imf_h(c(0, rev(MAPS[[2]]$maps[[1]]$seq.rf)))))
for(i in 1:5)
  b[[i]]$position<-cumsum(c(0,imf_h(MAPs.res[[i]]$maps[[1]]$seq.rf)))
plot(0, type = "n", 
     axes = FALSE, 
     ylim = c(0, max(res[c("polymapR.length", "MAPPoly.length"),])), 
     xlim = c(0, 17), 
     xlab = "Maps",
     ylab = "cM")
axis(2)
xid<-matrix(c(0,1,3,4,6,7,9,10,12,13), nrow = 2)
for(j in 1:5)
{
  lines(x = rep(xid[1,j], 2), y=range(a[[j]]$position))
  lines(x = rep(xid[2,j], 2), y=range(b[[j]]$position))
  idM<-match(a[[j]]$marker, b[[j]]$marker)
  idP<-match(b[[j]]$marker, a[[j]]$marker)
  points(x = rep(xid[1,j], nrow(a[[j]])), y=a[[j]]$position, pch = 19, col = scales::alpha(c("#e41a1c", "#377eb8")[is.na(idM)+1], .2) )
  points(x = rep(xid[2,j], nrow(b[[j]])), y=b[[j]]$position, pch = 19, col = scales::alpha(c("#e41a1c", "#377eb8")[is.na(idP)+1], .2))
  for(i in 1:nrow(a[[j]]))
  {
    if(!is.na(idM[i]))
      lines(c(xid[1,j],xid[2,j]), c(a[[j]]$position[i], b[[j]]$position[idM[i]]), lwd = .5)
  }
}
text(x = apply(xid,2,mean), y = rep(105, 5), labels = c("LG1","LG2","LG3","LG4","LG5"))
legend(14, 50, c("shared markers", "private markers"), col = c( "#e41a1c", "#377eb8"), pch = 19)
```

```{r, save_chunk, echo=FALSE, include=FALSE, eval=FALSE}
save.image(file = "~/repos/tutorials/polymapR_example/mapping_for_polymapR_example.rda", compress = TRUE)
```


